Bifurcation or state tipping: assessing transition type in a model trophic cascade

### 2018-12-14 paper/paper.Rmd
completed revisions

Central to resolving this question is untangling a second issue that frequently arises in the discussion of alternative stable state dynamics in ecological systems: the issue of dimension.   Ecosystems are notoriously complex, and this lake ecosystem is no exception. Here, we analyze a previously published model of a trophic cascade to determine whether manipulating the top predator constitutes B-tipping or S-tipping. Common versions of this model include 5 state variables, ranging from top predators to primary producers (Figure \ref{fig2}). In sharp contast to this complexity, both intutition and theory of critical transitions have been centered on one-dimensional intpretations. The S-tipping and B-tipping transitions illustrated in Figure \ref{fig1} are fundamentally rooted in a one-dimensional analysis: the state consists of a single continuously valued dynamical variable (i.e. population density of adult bass).  Similarly, Ball-in-cup style depictions of alternative stable states and sudden transitions present only a one-dimensional dipiction of a system state. While it may be tempting to dismiss such simple depictions as out-of-hand as mere cartoons, it is in fact possible to map the full five-dimensional model explicitly into a one-dimensional approximation under a few simple assumptions.  Here, we present the full dimension reduction on the full 5D model analytically, and compare this approximation to numerical simulations of the original system. This exercise facilitates the interpretation of the lake experiment in the context of predicting regime shifts by using a 1D model to understand a high dimensional system.

@@

This analysis also highlights the importance of what might appear to be a merely semantic issue that in fact corresponds to fundamentally different dynamics. In this formulation, the control parameter $q$ corresponds to harvest (positive $q$) or stocking (negative $q$) of adult bass in the lake. It is this aspect that first gives the appearance of state tipping dynamics, since it suggests the state is being somehow "directly manipulated". Yet from a mathematical standpoint, it is irrelevant if the control parameter corresponds to some environmental or policy variable (fishing regulations, fish migration into the lake), which in turn increases or decreases net mortality proportionally, or if the manipulation directly adds or removes adult bass -- the feature that distinguishes this manipulation from a S-tipping transition comes down to the scales involved. This can be seen most clearly though an analogy of considering the same model applied to a larger scale, such as marine fisheries.  In fisheries models, one would routinely examine the dynamics of the fish population biomass $B$ with respect to different possibly parameters for harvests or fishing effort.  Even though harvest may be measured in the same units as the total biomass, it is natural to think of the two quantites very differently.  A manager can adjust the harvest effort by changing quotas, limiting permits for fishing boats etc.  Slowly increasing in the amount of bass stocked in the lake (i.e. decreasing $q$, effectively running the scenario of increasing fishing harvests in reverse) gives the rest of the food web time to adjust.  In contrast, a single large introduction of many bass would create an S-tipping transition, driving the system immediately into an alternate state without exhibiting CSD phenomena needed to evaluate early warning signals of a tipping point.

@@

More generally, our analysis draws attention to issues of dimensionality that have been largely overlooked despite significant interest in questions of regime shifts and early warning signals [e.g. @Clements2018; @Scheffer2015; @Scheffer2012].  The popularity of this topic has sometimes led to casually equating any sudden shift with fold bifurcations, overlooking any assumptions necessary to approximate a complex dynamical system with such a one-dimensional model.  Influential reviews on abrupt change in ecological systems [e.g. @Ratajczak2018] focus almost exclusively on descriptions of one-dimensional state-spaces without discussion of if, when or how complex systems such as the ecosystem food-web considered here can be appropriately described by a nonlinear equation in a single dimension.  The analysis shown here illustrates this process on the basis of a fast-slow timescale approximation, which allows us to replace the faster dynamics at lower trophic levels with their equilibrium states.  (Note also that the nonlinearity necessary to support alternate stable states only emerges through this approximation.)  Thus, for the five-dimensional model considered here, our analysis shows that the familiar one-dimensional characterization of a sudden shift being driven by a fold bifurcation is indeed appropriate, but this was not altogether evident a priori.  The fact that CSD phenomena can occur in many different models has been suggested to mean that CSD is a generic or universal early warning indicator [@Scheffer2009] of a critical transition.  In practice, many higher dimensional models may not be so well approximated along a single manifold, or may correspond to other bifurcations or transitions [@TE2013].  As such, analyses of regime shifts should not be too quick to sweep these details under the rug.  Mechanistic understanding will always require models.

>>>>

Central to resolving this question is untangling a second issue that frequently arises in the discussion of alternative stable state dynamics in ecological systems: the issue of dimension.   Ecosystems are notoriously complex, and this lake ecosystem is no exception. Here, we analyze a previously published model of a trophic cascade to determine whether manipulating the top predator constitutes B-tipping or S-tipping. Common versions of this model include 5 state variables, ranging from top predators to primary producers (Figure \ref{fig2}). In sharp contrast to this complexity, both intuition and theory of critical transitions have been centered on one-dimensional interpretations. The S-tipping and B-tipping transitions illustrated in Figure \ref{fig1} are fundamentally rooted in a one-dimensional analysis: the state consists of a single continuously valued dynamical variable (i.e. population density of adult bass).  Similarly, Ball-in-cup style depictions of alternative stable states and sudden transitions present only a one-dimensional depiction of a system state. While it may be tempting to dismiss such simple depictions as out-of-hand as mere cartoons, it is in fact possible to map the full five-dimensional model explicitly into a one-dimensional approximation under a few simple assumptions.  Here, we present the full dimension reduction on the full 5D model analytically, and compare this approximation to numerical simulations of the original system. This exercise facilitates the interpretation of the lake experiment in the context of predicting regime shifts by using a 1D model to understand a high dimensional system.

@@

This analysis also highlights the importance of what might appear to be a merely semantic issue that in fact corresponds to fundamentally different dynamics. In this formulation, the control parameter $q$ corresponds to harvest (positive $q$) or stocking (negative $q$) of adult bass in the lake. It is this aspect that first gives the appearance of state tipping dynamics, since it suggests the state is being somehow "directly manipulated". Yet from a mathematical standpoint, it is irrelevant if the control parameter corresponds to some environmental or policy variable (fishing regulations, fish migration into the lake), which in turn increases or decreases net mortality proportionally, or if the manipulation directly adds or removes adult bass -- the feature that distinguishes this manipulation from a S-tipping transition comes down to the scales involved. This can be seen most clearly though an analogy of considering the same model applied to a larger scale, such as marine fisheries.  In fisheries models, one would routinely examine the dynamics of the fish population biomass $B$ with respect to different possibly parameters for harvests or fishing effort.  Even though harvest may be measured in the same units as the total biomass, it is natural to think of the two quantities very differently.  A manager can adjust the harvest effort by changing quotas, limiting permits for fishing boats etc.  Slowly increasing in the amount of bass stocked in the lake (i.e. decreasing $q$, effectively running the scenario of increasing fishing harvests in reverse) gives the rest of the food web time to adjust.  In contrast, a single large introduction of many bass would create an S-tipping transition, driving the system immediately into an alternate state without exhibiting CSD phenomena needed to evaluate early warning signals of a tipping point.

@@

A second issue highlighted by this analysis is the contrast between the interpretation of the five-dimensional food web model and the classic one-dimensional description of tipping points shown in Fig \ref{fig1}. Such one-dimensional descriptions of a state are frequently presented in reviews or abstract discussion of regime shifts and early warnings with little or no discussion of how a complex system comes to be represented by a one-dimensional continuous variable [e.g. @Scheffer2001; @Scheffer2009; @Barnosky2012; @Scheffer2012; @Clements2018], and even literature focusing on specific systems such as the lake-ecosystem  considered here can be vague about the dimensional reduction [@Carpenter2008; @Carpenter2011].  This can in turn create confusion as to whether such one-dimensional diagrams are meant to be interpreted as mathematically explicit approximations of higher-order dynamics, or merely as cartoons meant only to mimic the qualitative pattern but bearing no quantitative relationship with the actual process. The analysis presented here for the lake-ecosystem model demonstrates that the former does indeed hold, though only under the appropriate time-scale limits -- not only that of the "slow parameter" but also that of the other dynamic variables in the system: plankton, Juvenile Bass, Forage Fish, Fig \ref{fig2}. Our analysis also highlights that how the nonlinearity required to create a fold bifurcation arises directly from this dimension-reducing approximation.  In the original five-dimensional model, Eq (1), the dynamics of the adult bass $A$ appear to be completely linear.  Only after accounting for the feedback in the other dimensions do we see the nonlinearity required to create a fold bifurcation.  Our numerical simulations provide an illustration of how closely the dynamics of the full five-dimensional system are captured by the one-dimensional approximation using a biologically plausible parameterization. Thus the reference to the dynamics of a one-dimensional system state, as well as the interpretation of the transition as B-tipping rather than S-tipping, appears to be well-justified in the lake-ecosystem case.

This analysis does not vindicate the use of one-dimensional models more generally.  The popularity and simplicity of the fold-bifurcation model has sometimes led to more casually equating any sudden shift with fold bifurcations, overlooking any assumptions necessary to approximate a complex dynamical system with such a one-dimensional model.  Influential reviews on abrupt change in ecological systems [e.g. @Scheffer2015; @Ratajczak2018] focus almost exclusively on descriptions of one-dimensional state-spaces without discussion of if, when or how complex systems such as the ecosystem food-web considered here can be appropriately described by a nonlinear equation in a single dimension.  More careful modeling informed by system-specific knowledge is required before any abrupt transition can be characterized as a fold bifurcation [@Nature2013]. The fact that CSD phenomena can occur in many different models has been suggested to mean that CSD is a generic or universal early warning indicator [@Scheffer2009] of a critical transition.  In practice, many higher dimensional models may not be so well approximated along a single manifold, or may correspond to other bifurcations or transitions [@TE2013].  As such, analyses of regime shifts should not be too quick to sweep these details under the rug.  Mechanistic understanding will always require models.
